Red-Giant Hot Jupiters: Brilliant Radio Beacons?

### 2015-08-29 ms.tex
ready to read sensitivity section

\documentclass{emulateapj}

\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{rotating}
\usepackage{natbib}
%\usepackage{lscape}
\usepackage{enumerate}
\usepackage{multirow}
\usepackage{array}
\usepackage{appendix}
\usepackage{comment}
\usepackage{color}
\usepackage{url}
%\usepackage[dvipdfmx]{color}
%\usepackage[dvipdfmx]{graphicx}

@@

%%%%%%%%%% 
%\begin{eqnarray}
%B   &\sim &  10~{\rm G} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.38} \label{eq:scalingB} \\
%B   &\sim &  10~{\rm G} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.04} \label{eq:scalingB} \\
%f_{\rm cyc} &\sim &  28~{\rm MHz} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.38} \label{eq:scalingfc} 
%f_{\rm cyc} &\sim &  28~{\rm MHz} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.04} \label{eq:scalingfc} 
%\end{eqnarray}
%%%%%%%%%%
%under this model. 
Note that the cyclotron frequency of Jovian planets typically falls between 10~MHz, the plasma cut-off frequency of Earth's ionosphere, and less than 1~GHz. 
In this regime, there are a number of current and near-future radio wave detectors including GMRT, LOFAR, LWA, MWA, Hera, SKA, and upgrades to the VLA.

@@

\memoYF{discussion about detectability... Figures 1-3 of \citet{griesmeier2007b} show $10^{-3}$ Jy limit for 10-100 MHz (LWA, LOFAR) and $10^{-5}-10^{-4}$ Jy for 100-1000 MHz (LOFAR, SKA).}

We consider in this section several current and near future radio observatories that 
operate at frequencies $\nu \lesssim 200~\rm MHz$ and can reach continuum 
sensitivities less than 1~mJy in a few hours.  The results are summarized
in Table~\ref{tab:sens}.
The primary limitations to detectability are noise, characterized by the sky
and system equivalent blackbody temperatures $T_{\rm sky}$ and $T_{\rm sys}$,
resolution, and source confusion.  
To estimate the source confusion $\sigma_c$ (in mJy beam$^{-1}$) reported in Table~\ref{tab:sens}, 
we use equations provided by \cite{condon+ransom2016}, reproduced below:
\begin{eqnarray}
\sigma_c & \approx & 0.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^2\!\!\!\!,~(\theta > 10\arcsec) \\
\sigma_c & \approx & 2.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^{10/3}\!\!\!\!\!\!\!\!\!\!,~(\theta < 10\arcsec).
\end{eqnarray}
The source confusion limit is dominated by active galactic nuclei, which have low polarization 
fractions $\lesssim2.5\%$ \citep{Stil2014} and slowly vary with time \citep{Condon1974,Condon2012}.
Therefore both the circularly polarized nature of the emission from hot Jupiters 
and the time variability (discussed in \S\ref{ss:timevariability}) partially mitigate 
the source confusion limit.

The LOw-Frequency ARray (LOFAR) operates as two separate arrays defined by their 
high and low bands, differing configurations, and antenna designs.  
LOFAR-LBA operates in the 10-90~MHz range, and LOFAR-HBA operates in the 
100-240~MHz range \cite{vanHaarlem2013}.  We consider it as two separate
instruments here.
LOFAR-LBA's most sensitive band is centered at 60~MHz, and at 150~MHz 
for LOFAR-HBA.  Given the strong frequency dependence of
the effective aperture of a dipole antenna, and that the system noise temperature 
$T_{\rm sys}$ for LOFAR is not fully determined by the sky noise $T_{\rm sky}$, 
we use the LOFAR Image Noise calculator\footnote{
Heald; \url{http://www.astron.nl/~heald/test/sens.php}}, which 
is based on SKA Memo 113\footnote{
Nijboer, Pandey-Pommier, \& de Bruyn; 
\url{http://www.skatelescope.org/uploaded/59513\_113\_Memo\_Nijboer.pdf}}, for
our fiducial sensitivity and imaging parameters in Table~\ref{tab:sens}.

The Giant Metre-wave Radio Telescope (GMRT) is an array comprising thirty 
25-meter antennas, operating at frequencies 130-2000~MHz. The lowest
band (130-190~MHz) exhibits receiver noise comparable to the sky noise, and
hence we use their online calculator.\footnote{
\href{http://gmrt.ncra.tifr.res.in/~astrosupp/obs_setup/sensitivity.html}}

The low frequency component of the  Square Kilometer Array (SKA), recently rebaselined 
to have 0.4~km$^2$ collecting area in 130,000 elements with baselines extending
to 65~km, will operate from 50-350~MHz.\footnote{
\url{https://www.skatelescope.org/news/worlds-largest-radio-telescope-near-construction/}}
We estimate the noise in the lower half of the band ($\nu<200~\rm MHz$) assuming
$T_{\rm sys} \approx T_{\rm sky}$, and use the fit from \cite{Rogers+Bowman2008}:
\begin{equation}
T_{\rm sky} = T_{150} \left( \frac{\nu}{150~\rm MHz}\right)^{-\beta} + T_{\rm CMB} 
\end{equation}
where $T_{150} = 283.2~\rm K$ is the $T_{\rm sky}$ at 150~MHz, $\beta=2.47$,
and $T_{\rm CMB} = 2.73~\rm K$ is the contribution from the cosmic microwave 
background (CMB).
While this relation was fit to data taken in the 100-200~MHz range, the authors
found it to be consistent with published measurements from 10-408~MHz.

For the rest of the instruments considered, we also assume $T_{\rm sys} \approx T_{\rm sky}$,
and adopt the \cite{Rogers+Bowman2008} relation.  We caution that is assumption is 
optimistic, and the time estimates could in reality be a factor of a few worse.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table*}[htp]
\caption{Detectability of Hot Jupiters with Current and Future Radio Telescopes.}
\begin{center}
\begin{tabular}{c|ccccc} \hline \hline\\[-0.5pc]
Instrument & Band & Sensitivity & $\tau$  & Resolution & $5\sigma_c$\\
 & MHz & $\mu$Jy~hr$^{1/2}$ &  hr & \arcsec\ & $\mu$Jy bm$^{-1}$ \\[0.25pc] 
\hline\\[-0.5pc]
GMRT      & 130-190 & 430   &  460             & 16.5 & 280 \\ 
HERA      &  30-80  & 790   &  730             & 1030 & $2.1\times10^6$\\ 
LOFAR-LBA &  30-80  & 700   & 1225             & 8.3  & 211 \\ 
LOFAR-HBA & 110-200 &  35   &    3             & 3.3  & 5.5 \\  
LWA       &  30-80  & 53000 &  3.2$\times10^6$ & 10.3 & 211 \\
MWA       &  30-80  & 27000 &  8.1$\times10^5$ & 1030 & $2.1\times10^6$ \\
SKA-Low   & 50-350  &  4.9   &  0.03            & 7.9  & 57 \\
VLA       & 30-80   & 138  & 4760 & \\
\hline
\end{tabular}
\end{center}
\label{tab:sens}
\tablecomments{Time estimates to reach 20~$\mu$Jy RMS.}
\end{table*}

@@

>>>>

\documentclass[iop,numberedappendix,apj]{emulateapj}

\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{rotating}
\usepackage{natbib}
%\usepackage{lscape}
\usepackage{enumerate}
\usepackage{multirow}
\usepackage{array}
\usepackage{appendix}
\usepackage{comment}
\usepackage{color}
\usepackage{url}
\usepackage{hyperref}
%\usepackage[dvipdfmx]{color}
%\usepackage[dvipdfmx]{graphicx}

@@

%%%%%%%%%% 
%\begin{eqnarray}
%B   &\sim &  10~{\rm G} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.38} \label{eq:scalingB} \\
%B   &\sim &  10~{\rm G} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.04} \label{eq:scalingB} \\
%f_{\rm cyc} &\sim &  28~{\rm MHz} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.38} \label{eq:scalingfc} 
%f_{\rm cyc} &\sim &  28~{\rm MHz} \left( \frac{t}{4.5~\rm{Gyr}} \right)^{-0.43} \left( \frac{M_p }{M_{p,J}} \right)^{1.04} \label{eq:scalingfc} 
%\end{eqnarray}
%%%%%%%%%%
%under this model. 
Note that the cyclotron frequency of Jovian planets typically falls between 10~MHz, the plasma cut-off frequency of Earth's ionosphere, and less than 1~GHz. 
In this regime, there are a number of current and near-future radio wave detectors including GMRT, 
LOFAR, Hera, SKA, and upgrades to the VLA.

@@

%\memoYF{discussion about detectability... Figures 1-3 of \citet{griesmeier2007b} show $10^{-3}$ Jy limit for 10-100 MHz (LWA, LOFAR) and $10^{-5}-10^{-4}$ Jy for 100-1000 MHz (LOFAR, SKA).}

We consider in this section several current and near future radio observatories that 
operate at frequencies $\nu \lesssim 350~\rm MHz$ and can reach continuum 
sensitivities less than 1~mJy in a few hours.  The results are summarized
in Table~\ref{tab:sens}.
The primary limitations to detectability are noise, characterized by the
system equivalent blackbody temperature $T_{\rm sys}$, resolution, and source confusion.  
To estimate the source confusion $\sigma_c$ (in mJy beam$^{-1}$) reported in Table~\ref{tab:sens}, 
we use the relations provided by \cite{condon+ransom2016}, reproduced below:
\begin{eqnarray}
\sigma_c & \approx & 0.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^2\!\!\!\!,~(\theta > 10\arcsec) \\
\sigma_c & \approx & 2.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^{10/3}\!\!\!\!\!\!\!\!\!\!,~(\theta < 10\arcsec).
\end{eqnarray}
The source confusion limit is dominated by active galactic nuclei, which have low polarization 
fractions $\lesssim2.5\%$ \citep{Stil2014} and slowly vary with time \citep{Condon1974,Condon2012}.
Therefore both the circularly polarized nature of the emission from hot Jupiters 
and the time variability (discussed in \S\ref{ss:timevariability}) partially mitigate 
the source confusion limit, allowing hot Jupiters, like other time-variable sources,
to be detected below the imaging confusion limit.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table*}[htp]
\caption{Detectability of Hot Jupiters with Current and Future Radio Telescopes.}
\begin{center}
\begin{tabular}{c|ccccc} \hline \hline\\[-0.5pc]
Instrument & Band & Sensitivity & $\tau$  & Resolution & $5\sigma_c$\\
 & MHz & $\mu$Jy~hr$^{1/2}$ &  hr & \arcsec\ & $\mu$Jy bm$^{-1}$ \\[0.25pc] 
\hline\\[-0.5pc]
GMRT       & 130-190 & 430   &  460             & 16.5 & 280 \\ 
HERA       &  30-80  & 790   &  870             & 1030 & $2.1\times10^6$\\ 
LOFAR-LBA  &  30-80  & 700   & 1225             & 8.3  & 210 \\ 
LOFAR-HBA  & 110-200 &  35   &    3             & 3.3  & 5.5 \\  
%LWA        &  30-80  & 53000 &  3.2$\times10^6$ & 10.3 & 211 \\
%MWA        &  30-80  & 27000 &  8.1$\times10^5$ & 1030 & $2.1\times10^6$ \\
SKA-Low (part) & 50-200 & 7.8 &  0.15           & 7.8  & 48 \\
SKA-Low (full) & 50-350 & 2.2 &  0.012          & 2.2  & 7.2 \\
%VLA 4-Band & 30-80   & 3300  & 1100             & 28.6 & 1600 \\
VLA-LOBO   &  50-350 &  92   &  21              & 8.6  & 52 \\
\hline
\end{tabular}
\end{center}
\label{tab:sens}
\tablecomments{Time estimates to reach 20~$\mu$Jy RMS.}
\end{table*}

@@

The LOw-Frequency ARray (LOFAR) operates as two separate arrays defined by their 
high and low bands, differing configurations, and antenna designs, which
are called respectively the Low Band Antennas (LBA) and High Band Antennas (HBA).  
LOFAR-LBA operates in the 10-90~MHz range, and LOFAR-HBA operates in the 
100-240~MHz range \citep{vanHaarlem2013}.  We consider it as two separate
instruments here.
LOFAR-LBA's most sensitive band is centered at 60~MHz, and LOFAR-HBA's is at 150~MHz. 
Given the strong frequency dependence of
the effective aperture of a dipole antenna, and that the system noise temperature 
$T_{\rm sys}$ for LOFAR is not fully determined by the sky noise $T_{\rm sky}$, 
we use the LOFAR Image Noise calculator\footnote{
Heald; \url{http://www.astron.nl/~heald/test/sens.php}}, which 
is based on SKA Memo 113\footnote{
Nijboer, Pandey-Pommier, \& de Bruyn; 
\url{http://www.skatelescope.org/uploaded/59513\_113\_Memo\_Nijboer.pdf}}, for
our fiducial sensitivity and imaging parameters in Table~\ref{tab:sens}.

The Giant Metre-wave Radio Telescope (GMRT) is an array comprising thirty 
25-meter antennas, operating at frequencies 130-2000~MHz. The lowest
band (130-190~MHz) exhibits receiver noise $T_{\rm rx}$ comparable to the sky noise.
To be conservative, we use the GMRT online calculator.\footnote{
\url{http://gmrt.ncra.tifr.res.in/~astrosupp/obs\_setup/sensitivity.html}
}

For the low frequency receivers (28-80~MHz; 4-Band) in the % Long Wavelength Array (LWA) and the 
Very Large Array (VLA), the receiver noise temperature $T_{\rm rx}\approx 260~\rm K$ 
\citep{Hicks2012} and is subdominant to $T_{\rm sky}$ by at least -6~dB.  Future
upgrades to the VLA such as the LOw Band Observatory \citep[LOBO][]{Kassim2015IAU},
We estimate the sensitivity of LOBO assuming the receiver noise extrapolates linearly
from $T_{\rm rx}\approx 60~\rm K$ (internal VLA memo) in P-Band (230-470~MHz) to
$T_{\rm rx}\approx 260~\rm K$ in 4-band, and consider just the 50-350~MHz portion
of the full 30-470~MHz band.

We compute $T_{\rm sys} = T_{\rm rx}+ T_{\rm sky}$ using the low frequency sky 
noise temperature fit from \cite{Rogers+Bowman2008}:
\begin{equation}
T_{\rm sky} = T_{150} \left( \frac{\nu}{150~\rm MHz}\right)^{-\beta} + T_{\rm CMB} 
\end{equation}
where $T_{150} = 283.2~\rm K$ is the $T_{\rm sky}$ at 150~MHz, $\beta=2.47$,
and $T_{\rm CMB} = 2.73~\rm K$ is the contribution from the cosmic microwave 
background (CMB).
While this relation was fit to data taken in the 100-200~MHz range, the authors
found it to be consistent with published measurements from 10-408~MHz.

The low frequency component of the  Square Kilometer Array (SKA), recently rebaselined 
to have 0.4~km$^2$ collecting area in 130,000 elements with baselines extending
to 65~km, will operate from 50-350~MHz.\footnote{
\url{https://www.skatelescope.org/news/worlds-largest-radio-telescope-near-construction/}}
We estimate the noise in the lower half of the band ($\nu<200~\rm MHz$), appropriate
for lower mass ($\lesssim 10 M_J$) separately from the full band, which could be used
to detect $\sim 30 M_J$ hot Jupiters.
We adopt the \cite{Rogers+Bowman2008} relation for $T_{\rm sys}$ as above, and assume
$T_{\rm rx}\approx 60~\rm K$.
We caution that these assumptions are optimistic, and the time estimates could in reality 
be a factor of a few worse.
